Faint Object Detection in Multi-Epoch Observations\\ Via Interim Catalog Data Fusion

### 2016-11-05 ms.tex
Title change: Interim catalog data fusion
Revised fig captions, added astrometric BF fig (f7) and text, renumbered last few fig file references

\title{Incremental Faint Object Detection in Multi-Epoch Observations}
%\journalinfo{To appear in ApJ?}
\journalinfo{Submitted to Apj}

@@

\begin{abstract} 
Observational astronomy in the time-domain era faces several new challenges. 
One of them is the efficient use of observations obtained at multiple epochs. 
The work presented here addresses faint object detection with multi-epoch data, and describes an incremental strategy for separating real objects from artifacts in ongoing surveys, in situations where the single-epoch data are summaries of the full image data, such as single-epoch catalogs of flux and direction estimates (with uncertainties) for candidate sources.
The basic idea is to produce low-threshold single-epoch catalogs, and use a probabilistic approach to accumulate catalog information across epochs; this is in contrast to more conventional strategies based on co-added or stacked image data across all epochs.
We adopt a Bayesian approach, addressing object detection by calculating the marginal likelihoods for hypotheses asserting there is no object, or one object, in a small image patch containing at most one cataloged source at each epoch.
The object-present hypothesis interprets the sources in a patch at different epochs as arising from a genuine object; the no-object (noise) hypothesis interprets candidate sources as spurious, arising from noise peaks.
We study the detection probability for constant-flux objects in a simplified Gaussian noise setting, comparing results based on single exposures and stacked exposures to results based on a series of single-epoch catalog summaries.
Computing the detection probability based on catalog data amounts to generalized cross-matching: it is the product of a factor accounting for matching of the estimated fluxes of candidate sources, and a factor accounting for matching of their estimated directions (i.e., directional cross-matching across catalogs).
We find that probabilistic fusion of multi-epoch catalog information can detect sources with only modest sacrifice in sensitivity compared to stacking.
The probabilistic cross-matching framework underlying our approach plays an important role in maintaining detection sensitivity, and points toward generalizations that could accomodate variability and complex object structure.
\end{abstract}

@@

We first analyze the simulated data considering only the photometric information (i.e., ignoring the directional Bayes factors).
In Figure~\ref{fig:bf} the (red) points represent the resulting Bayes factors for the real sources (right of the double dashed vertical lines) and the noise peaks (on the left).
%
The weight of evidence is strong for the bright galaxies but weakens for the faint galaxies.
The smallest Bayes factors arise for galaxies with true magnitudes near 26.5, which corresponds to the mode of the noise peak distribution, $\npd(\fest)$.
Perhaps surprisingly, sources dimmer than magnitude 26.5 can have larger Bayes factors than those with magnitude 26.5.
This happens because $\npd(\fest)$ peaks away from $\fest=0$, i.e., we do not expect noise peaks to have arbitrarily small measured fluxes.
For the weakest detectable sources, the most likely number of detections among the $k=9$ epochs is one.
The flat top of the distribution at the faint end corresponds to very dim sources detected only once, very near threshold.
The smaller Bayes factors in that region of the plot correspond to unlikely larger numbers of detections near the threshold; the discreteness in the number of detections produces a subtle banding in the distribution.

@@

Figure~\ref{fig:bf2} shows the distribution of Bayes factors, for the real (mock) and noise sources, now including the astrometric cross-matching factor.
Banding due to discreteness in the number of detections is now clearly apparent among the true-object Bayes factors; the 9 levels correspond to the different number of detections with the lowest being 1.
For the lowest band, corresponding to a single detection, $B_{\rm{}pos}$ is unity by definition (no constraint coming from a single detection), producing the same Bayes factor distribution as in the flux-only calculation.
For multiple detections, the Bayes factors for the true sources are greatly enhanced by including astrometric information (note that the ordinate is logarithmic); just two detections produces quite strong evidence for the true-object hypothesis.
The Bayes factors for the noise peaks have moved to much lower values, due to the low likelihood of directional coincidences.

@@

%****************************************
\color{red}
\begin{figure}
\epsscale{1.2}
\plotone{fig/f7u-U.png}
\caption{The (red) points indicate the weight of evidence for simulated galaxies as a function of their true brightness and the same for generated random noise peaks.
The Bayes factor is high for the bright galaxies but its logarithm is close to zero for the faint ones.
The noise peaks (left of double lines) seem to separate slightly from the faintest sources (just right of the double dashed vertical lines) because the fluxes in noise peaks tend to be higher, which better mimics sources between 26 and 25 magnitudes.}
\label{fig:bf}
\end{figure}

\begin{figure}[t!]
\epsscale{1.2}
\plotone{fig/f8u-U.png}
\caption{The Bayes factor can differentiate significantly better between real objects and noise peaks when we include astrometric information.
Here we plot logarithm of product of the photometric and astrometric Bayes factors, the latter which comes directly from the cross-identification based on the celestial coordinates, see text for details.}
\label{fig:bf2}
\end{figure}

>>>>

\title{Faint Object Detection in Multi-Epoch Observations\\ Via Interim Catalog Data Fusion}
%\journalinfo{To appear in ApJ?}
\journalinfo{Submitted to Apj}

@@

\begin{abstract} 
Observational astronomy in the time-domain era faces several new challenges. 
One of them is the efficient use of observations obtained at multiple epochs. 
The work presented here addresses faint object detection with multi-epoch data, and describes an incremental strategy for separating real objects from artifacts in ongoing surveys, in situations where the single-epoch data are summaries of the full image data, such as single-epoch catalogs of flux and direction estimates (with uncertainties) for candidate sources.
The basic idea is to produce low-threshold single-epoch catalogs, and use a probabilistic approach to accumulate catalog information across epochs; this is in contrast to more conventional strategies based on co-added or stacked image data across all epochs.
We adopt a Bayesian approach, addressing object detection by calculating the marginal likelihoods for hypotheses asserting there is no object, or one object, in a small image patch containing at most one cataloged source at each epoch.
The object-present hypothesis interprets the sources in a patch at different epochs as arising from a genuine object; the no-object (noise) hypothesis interprets candidate sources as spurious, arising from noise peaks.
We study the detection probability for constant-flux objects in a simplified Gaussian noise setting, comparing results based on single exposures and stacked exposures to results based on a series of single-epoch catalog summaries.
Computing the detection probability based on catalog data amounts to generalized cross-matching: it is the product of a factor accounting for matching of the estimated fluxes of candidate sources, and a factor accounting for matching of their estimated directions (i.e., directional cross-matching across catalogs).
We find that probabilistic fusion of multi-epoch catalog information can detect sources with only modest sacrifice in sensitivity and selectivity compared to stacking.
The probabilistic cross-matching framework underlying our approach plays an important role in maintaining detection sensitivity, and points toward generalizations that could accomodate variability and complex object structure.
\end{abstract}

@@

We first analyze the simulated data considering only the photometric information (i.e., ignoring the directional Bayes factors).
In Figure~\ref{fig:bf-photo} the (red) points represent the resulting Bayes factors for the real sources (right of the double dashed vertical lines) and the noise peaks (on the left).
%
The weight of evidence is strong for the bright galaxies but weakens for the faint galaxies.
The smallest Bayes factors arise for galaxies with true magnitudes near 26.5, which corresponds to the mode of the noise peak distribution, $\npd(\fest)$.
Perhaps surprisingly, sources dimmer than magnitude 26.5 can have larger Bayes factors than those with magnitude 26.5.
This happens because $\npd(\fest)$ peaks away from $\fest=0$, i.e., we do not expect noise peaks to have arbitrarily small measured fluxes.
For the weakest detectable sources, the most likely number of detections among the $k=9$ epochs is one.
The flat top of the distribution at the faint end corresponds to very dim sources detected only once, very near threshold.
The smaller Bayes factors in that region of the plot correspond to unlikely larger numbers of detections near the threshold; the discreteness in the number of detections produces a subtle banding in the distribution.

@@

Figure~\ref{fig:bf-astro} shows the distribution of astrometric Bayes factors, for the real (mock) and noise sources.
Note the larger span of the (log) Bayes factor axis.
Comparing to Figure~\ref{fig:bf-photo}, we see that directional cross-matching is a stronger discriminant between real and noise sources in the dim source regime.
The photometric data grow in importance as sources grow brighter.

Figure~\ref{fig:bf-both} shows the distribution of Bayes factors, for the real and noise sources, now combining the photometric and astrometric factors.
Banding due to discreteness in the number of detections is now clearly apparent among the true-object Bayes factors; the 9 levels correspond to the different number of detections with the lowest being 1.
For the lowest band, corresponding to a single detection, $B_{\rm{}pos}$ is unity by definition (no constraint coming from a single detection), producing the same Bayes factor distribution as in the flux-only calculation.
For multiple detections, the Bayes factors for the true sources are greatly enhanced by including astrometric information (note that the ordinate is logarithmic); just two detections produces quite strong evidence for the true-object hypothesis.
The Bayes factors for the noise peaks have moved to much lower values, due to the low likelihood of directional coincidences.

@@

%****************************************
\color{red}
\begin{figure}
\epsscale{1.2}
\plotone{fig/f6.png}
\caption{Bayes factor distributions for simulated galaxies as a function of their true brightness (right axes), and the same for generated random noise peaks (left axes), using simulated photometric data (ignoring astrometric information).
Points (red) indicate the weight of evidence, quantified via (logarithmic) Bayes factors comparing real-source and noise origins for the simulated photometric data.
The Bayes factor is high for the bright galaxies but is close to unity (logarithm close to zero) for the faint ones.
The Bayes factor distribution for the noise peaks (left axes) peaks below the distribution for the faintest real sources (left region of right axes) because the measured fluxes in noise peaks tend to be in a range that mimics sources between 26 and 25 magnitudes, rather than the dimmest sources.}
\label{fig:bf-photo}
\end{figure}

\begin{figure}[t!]
\epsscale{1.2}
\plotone{fig/f7.png}
\caption{Bayes factor distributions as in Figure~\ref{fig:bf-photo}, but here accounting for only astrometric information, i.e., cross-identification across epochs based on the celestial coordinates estimates and uncertainties.
Note the larger range for the Bayes factor axis.}
\label{fig:bf-astro}
\end{figure}

\begin{figure}[t!]
\epsscale{1.2}
\plotone{fig/f8.png}
\caption{Bayes factor distributions as in Figure~\ref{fig:bf-photo}, but accounting for both photometric and astrometric information.
Points (red) indicate the logarithm of the product of the photometric and astrometric Bayes factors.
The astrometric factor dominates for weak sources, and the photometric factor dominates for bright ones.}
\label{fig:bf-both}
\end{figure}
